Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-26 paper.tex
more edits ot the paper

The model can also be tested in a slightly different approach by checking whether it can correctly
identify cliffs between a new molecule and the member of the training set. In fact, such a scenario
would be the anticipated, prospective use case for such a modeling methodology. To test this, we
consider the original set of $n$ individual molecules in a dataset and hold out $m$ molecules (as
opposed to holding out a set of observations from the pairwise dataset). We then evaluate the SALI
values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest model. In contrast to the
models previously described, we rebuilt these models using the entire set of $n$ molecules. The
model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
remaining molecules.

We first performed this analysis with the Kalla dataset. Since predictions of new molecules involve
$n$ predictions per molecule, we removed three molecules from the dataset and rebuilt the random
forest model. The hold out molecules, were specifically selected since they displayed significant
activity cliffs with various members of the training set. We then evaluated the pairwise descriptor
values for these three with the training set molecules and obtained predictions of the log(SALI)
values. Figure \ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training set
as well as for the prediction set. From a numerical point of view, the performance of the model on
he prediction set degrades somewhat (RMSE of 0.51 versus 0.41 for the training set). As noted above,
the variance of the prediction is higher at the lower scale of log(SALI) values. However, at the
higher end of the scale there a number of pairs that are well predicted. Yet, it is clear that a
number of relatively significant activity cliffs have been underestimated.

@@

Figure \ref{fig:kalla-comparison} displays the structures of the three hold out molecules (top row)
and a selection of structures, which which these hold out molecules were predicted to exhibit an
activity cliff. For each molecule, the $K_i$ value is listed along with the absolute residual value
(log(SALI) units). For molecules \textbf{349288} and \textbf{349699}, the predicted cliffs are
relatively accurate, as evidenced by the low residuals. For \textbf{349138} on the other hand, most
of the residuals were were relatively high. While we have highlighted a number of cliffs,
Figure \ref{fig:tset-pset-kalla}C makes it clear that the predictions for this molecule were, on
average, worse than for the other two molecules.

Figure \ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
applied to the Dai dataset. As before we removed three molecules (top row of
Figure \ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining
41 molecules. We then predicted the pairwise SALI values for the three hold-out molecules with the
remaining 41 molecules (i.e., the prediction set). A plot of the predicted versus observed log(SALI)
values for the training and prediction sets are displayed in Figure \ref{fig:tset-pset-dai}A. As
before the model performs poorly on the lesser cliffs, but improves for the more significant cliffs
(RMSE for the training set and prediction sets were 0.37 and 0.43 log units respectively). However,
the 2 most significant cliffs are under-estimated. However, in comparison to the Kalla dataset, the
distribution of absolute residuals aggregated by the hold-out molecule
(Figure \ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than
0.3 log units.  For this dataset, the number of ``real'' cliffs is relatively low as shown in
Figure \ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to
significant cliffs are relatively well predicted - in fact, the maximum residuals are observed for
the least significant cliffs.

We then considered the three hold-out molecules and some of the members of the training set with
which they are predicted to show activity cliffs. For this dataset, there are relatively few
``true'' activity cliffs. For example, while molecules \textbf{371259} and \textbf{371307} have a
1000-fold difference in activity, their $T_c = 0.68$, and would not represent a truly significant
cliff. However, the pair is well predicted with a residual of 0.04 log units. if we consider the
subset of predictions where the $T_c \geq 0.8$ we note that the median absolute residual is 0.24 log
units. However, the activity differences are not always significant ranging from 1.1-fold to
13-fold. Two such examples are shown in Figure \ref{fig:dai-comparison}, where the only difference is
in the position of the methyl substituent. Similar behavior is observed with the other hold-out
molecules.

@@

>>>>

The model can also be tested in a slightly different approach by checking whether it can correctly
identify cliffs between a new molecule and the members of the training set. In fact, such a scenario
would be the anticipated, prospective use case for such a modeling methodology. To test this, we
consider the original set of $n$ individual molecules in a dataset and hold out $m$ molecules (as
opposed to holding out a set of observations from the pairwise dataset). We then evaluate the SALI
values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest model. In contrast to the
models previously described, we rebuilt these models using the entire set of $n$ molecules. The
model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
remaining molecules. As a result, instead of a just $m$ predicted values, we obtain $m \times (n-m)$
predicted values for the test set.

We first performed this analysis with the Kalla dataset. Since predictions of new molecules involve
$n$ predictions per molecule, we removed ten molecules from the original dataset and rebuilt the
random forest model. The hold out molecules included cases that displayed significant activity
cliffs with various members of the training set, as well as other cases that did not really exhibit
cliffs with any member of the training set. We then evaluated the pairwise descriptor values for
these ten molecules with the training set molecules and obtained predictions of the log(SALI)
values. Figure \ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training
set as well as for the prediction set. From a numerical point of view, the performance of the model
on the prediction set degrades somewhat (RMSE of 0.41 versus 0.63 for the training set). As noted
above, the variance of the prediction is higher at the lower scale of log(SALI) values. This is
highlighted in Figure \ref{fig:sali-pred-variance}, where we divided the observed log SALI values
into three groups, \emph{low}, \emph{medium} and \emph{high} (based on the quartiles of the values)
and aggregated the prediction residuals for the observations belonging to each group. Clearly, for
lower SALI values (i.e., less significant cliffs), the prediction errors are larger and more widely
distributed. While \ref{fig:tset-pset-kalla}A indicates that a number of the more significant
activity cliffs are quite well predicted, Figure \ref{fig:sali-pred-variance} indicates that as a
group the observations with higher log SALI values tend to exhibit a small range of non-zero
residual errors. However it is clear from the scatter plot that number of relatively significant
activity cliffs have been underestimated.

@@

Figure \ref{fig:kalla-comparison} displays the structures of three hold out molecules (top row)
and a selection of structures, which which these hold out molecules were predicted to exhibit an
activity cliff. For each molecule, the $K_i$ value is listed along with the absolute residual value
(log(SALI) units). For molecules \textbf{349288} and \textbf{349699}, the predicted cliffs are
relatively accurate, as evidenced by the low residuals. For \textbf{349138} on the other hand, most
of the residuals were were relatively high. While we have highlighted a number of cliffs,
Figure \ref{fig:tset-pset-kalla}C makes it clear that the predictions for this molecule were, on
average, worse than for the other two molecules.

Figure \ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
applied to the Dai dataset. As before we removed ten molecules (top row of Figure
\ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining 34
molecules. We then predicted the pairwise SALI values for the ten hold-out molecules with the
remaining 34 molecules. A plot of the predicted versus observed log(SALI) values for the training
and prediction sets are displayed in Figure \ref{fig:tset-pset-dai}A. As before the model performs
poorly on the lesser cliffs, but improves for the more significant cliffs (RMSE for the training set
and prediction sets were 0.37 and 0.43 log units respectively). Figure \ref{fig:sali-pred-variance}
also highlights the fact that the model performs poorly at the lower end of the log(SALI)
spectrum. However, the 2 most significant cliffs are under-estimated. However, in comparison to the
Kalla dataset, the distribution of absolute residuals aggregated by the hold-out molecule (Figure
\ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than 0.3 log units.
For this dataset, the number of ``real'' cliffs is relatively low as shown in Figure
\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to significant
cliffs are relatively well predicted - in fact, the maximum residuals are observed for the least
significant cliffs.

We then considered three of hold-out molecules and some of the members of the training set with
which they are predicted to show activity cliffs. For this dataset, there are relatively few
``true'' activity cliffs. For example, while molecules \textbf{371259} and \textbf{371307} have a
1000-fold difference in activity, their $T_c = 0.68$, and would not represent a truly significant
cliff. However, the pair is well predicted with a residual of 0.04 log units. if we consider the
subset of predictions where the $T_c \geq 0.8$ we note that the median absolute residual is 0.24 log
units. However, the activity differences are not always very large ranging from 1.1-fold to
13-fold. Two such examples are shown in Figure \ref{fig:dai-comparison}, where the only difference
is in the position of the methyl substituent (\textbf{371259} and \textbf{371224}). Similar behavior
is observed with the other hold-out molecules.

@@

\ctable[cap={}, caption={A summary of the prediction residuals for log(SALI) values, grouped by
  whether the actual log(SALI) value for that observation was \emph{low}, \emph{medium} or
  \emph{high}. The grouping is based on  the quartiles of the observed values.}, figure, botcap,
label={fig:sali-pred-variance}]    
{cc}
{}
{\textbf{Kalla} & \textbf{Dai}  \NN
\includegraphics[width=0.5\linewidth]{sali-pred-variance-kalla} & 
\includegraphics[width=0.5\linewidth]{sali-pred-variance-dai}
}
